---
title: "02_segment_trees"
author: "Matthew Coghill"
date: "2023-01-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In the previous script, we created cleaned, classified, and normalized LAS tiles. In this script, we will locate trees using either individual tree detection, individual tree segmentation, or both. The heights of the located trees will be returned and used in the next script to model tree DBH and volume.

First, load the required packages. I am including my `lidR.li2012enhancement` package here because it contains the function `lmfxauto()` for finding tree heights using an automatically calculated search window.

```{r Packages, warning=FALSE, message=FALSE}

ls <- c("tidyverse", "lidR", "lidR.li2012enhancement", "sf", "future", "terra")
invisible(lapply(ls, library, character.only = TRUE))
rm(ls)

```

For custom functions in this script, we just need the one that defines our parallel environment when we are working with nested parallelism:

```{r Functions}

source("./99_other/lidr_cores.R")
lidr_threads <- core_use()$lidr_threads
future_threads <- core_use()$future_threads

```

There are multiple algorithms for individual tree detection (ITD) and individual tree segmentation (ITS). ITD returns POINT features identifying tree tops and contains data on tree height. ITS returns a point cloud with an extra column `treeID`, which identifies unique trees in the point cloud. The function `crown_metrics()` is then able to be used on the segmented point cloud to generate a variety of metrics, including tree height, number of points, crown area, etc., as well as return a POLYGON shape of the tree. ITD is faster to perform than ITS and, with the help of the `lmfxauto()` function from the `lidR.li2012enhancement` package, easier to understand.

For ITD, the general way of identifying tree tops involves defining a "search radius" or "window size" for each point to find a local maximum (this is generally referred to as a local maximum filter, or LMF, and is described well with the `lmf()` function). At each point in the point cloud, the algorithm will search within the 2D window size to determine whether the point is the highest point within that search radius; if it is, it is defined as a tree top. How do we determine what the best window size is? Well...we don't! We use the `lmfxauto()` algorithm to determine the optimal parameters to pass to the `lmf()` function.

For ITS, there are multiple algorithms to consider for segmenting the point cloud. The one that makes the most sense to me is the `li2012()` algorithm, where the only parameter to consider is the LiDAR point cloud. Other algorithms, including `watershed()`, `dalponte2016()`, and `silva2016()` require a canopy height model or trees defined from an ITD algorithm. `li2012()`, on the other hand, is much simpler to navigate and we can pass the maximum tree radius found in the first script to this algorithm to speed up the computation speed. I am working on an offshoot of the `li2012()` algorithm, however it is not ready to be used in this instance yet so I will avoid using it in any significant capacity here.

In the chunk below, we will perform either ITD, ITS, or both depending on the switches we define:

`itd <- TRUE` indicates that ITD using the `lmfxauto()` algorithm will be performed
`its <- TRUE` indicates that ITS using the `li2012()` algorithm will be performed

If a switch is set to `FALSE` then those functions will not be run. If both are set to `FALSE` then an error should occur.

```{r Detect and Segment Trees}

# Define input and output directories, and create them where necessary
norm_dir <- file.path("./03_clean_norm")
norm_trees_dir <- file.path("./04_clean_norm_tree")
clean_trees_dir <- file.path("./05_clean_tree")
tree_shp_dir <- file.path("./06_tree_shp")
dir.create(clean_trees_dir, showWarnings = FALSE)
dir.create(norm_trees_dir, showWarnings = FALSE)
dir.create(tree_shp_dir, showWarnings = FALSE)
blocks <- dir(norm_dir)

# Define minimum height of what a tree should be, and max crown radius
th_tree <- 12
tree_rad <- as.numeric(readLines(file.path(tree_shp_dir, "tree_radius.txt")))

# itd = perform individual tree detection (returns points of treetop locations)
# its = individual tree segmentation (returns segmented point cloud and shapes 
# of trees) - this will take a long time in comparison to itd
itd <- TRUE
its <- FALSE

# Iterate ITD/ITS through each of the blocks
trees <- lapply(blocks, function(x) {
  
  # Create output directories for tiles with trees identified that are
  # normalized, as well as non-normalized tiles.
  tree_las_out <- file.path(norm_trees_dir, x)
  tree_las_unnorm_out <- file.path(clean_trees_dir, x)
  dir.create(tree_las_out, showWarnings = FALSE)
  dir.create(tree_las_unnorm_out, showWarnings = FALSE)
  
  # Read the LAS catalog
  block <- readUAVLAScatalog(file.path(norm_dir, x))
  
  if(itd) {
    
    # Prepare parallel environment for individual tree detection
    set_lidr_threads(lidr_threads)
    plan(multisession, workers = future_threads)
    
    # Set catalog processing options
    opt_chunk_buffer(block) <- 12.5
    opt_filter(block) <- "-drop_z_below 0"
    opt_output_files(block) <- ""
    
    # Run ITD. Sometimes, the algorithm produces a list object instead of an
    # `sf` dataframe, so coerce to `sf` dataframe in those cases
    ttops <- locate_trees(block, lmfxauto(), uniqueness = "bitmerge") 
    if(!is(ttops, "sf")) ttops <- do.call(rbind, ttops)
    
    # Add information regarding the block_year from the "x" variable
    ttops <- ttops |> 
      mutate(block = x) |> 
      dplyr::relocate(block, 1) |>  
      dplyr::relocate(geometry, .after = last_col())
    
    # Write the shape to a geopackage POINT file
    st_write(ttops, file.path(tree_shp_dir, paste0(x, "_treetops.gpkg")), 
             quiet = TRUE, delete_dsn = TRUE)
    
  } else if(!its) stop("At least one of 'itd' or 'its' must be TRUE")
  
  if(its) {
    
    # Prepare parallel environment for individual tree segmentation
    set_lidr_threads(1L)
    plan(multisession, workers = min(availableCores(), nrow(block)))
    
    # Set catalog processing options
    opt_chunk_buffer(block) <- 12.5
    opt_filter(block) <- "-drop_z_below 0 -drop_class 2"
    opt_output_files(block) <- file.path(tree_las_out, "{*}_trees")
    
    # Run ITS, and change the `future` plan to `sequential` to close the 
    # processes and free up RAM
    trees <- segment_trees(block, li2012(hmin = th_tree, speed_up = tree_rad))
    plan(sequential)
    set_lidr_threads(0L)
    
    # In the above process, ground points were filtered out before processing
    # to increase the processing speed of the tree segmentation algorithm and 
    # because it doesn't make sense for ground points to be labelled as a tree.
    # Here, we need to rebind the tree points and ground points together. We can 
    # accomplish that by loading separate LAS objects and then using rbindlist()
    block_files <- block$filename
    las_combine <- sapply(block_files, function(i) {
      
      # Identify the associated output tile with segmented trees from the 
      # original tile
      tree_file <- trees$filename[
        which(startsWith(
          substr(basename(trees$filename), 1, nchar(basename(trees$filename)) - 4),
          substr(basename(i), 1, nchar(basename(i)) - 4)))]
      
      if(length(tree_file)) {
        
        # If trees were segmented, bind the segmented tree file with the ground
        # classified points in the tile it originated from
        las_trees <- readLAS(tree_file)
        las_grnd <- readLAS(i, filter = "-drop_z_below 0 -keep_class 2")
        las_trees@data <- data.table::rbindlist(list(
          las_trees@data, las_grnd@data), fill = TRUE)
      } else {
        
        # If no trees were segmented, load the tile with all points
        las_trees <- readLAS(i, filter = "-drop_z_below 0")
      }
      
      # Write the bound tile with an associated .lax file
      writeLAS(las_trees, file.path(tree_las_out, basename(i)), index = TRUE)
      
      # Unnormalize the heights of the LAS file (returns to true heights), and 
      # write to a LAS file
      las_unnorm <- unnormalize_height(las_trees)
      writeLAS(las_unnorm, file.path(tree_las_unnorm_out, basename(i)), index = TRUE)
      return(file.path(tree_las_out, basename(i)))
    })
    
    # Next, we need to create the polygons of the individual trees in the 
    # catalog and gather their associated metrics. First, read the LAS catalog
    # and set the catalog processing options
    las_combine <- readUAVLAScatalog(tree_las_out)
    opt_chunk_buffer(las_combine) <- 12.5
    opt_output_files(las_combine) <- ""
    
    # Set parallel environment
    set_lidr_threads(1L)
    plan(multisession, workers = min(
      round(0.8 * availableCores()), nrow(block)))
    
    # Generate tree shapes and other standard crown metrics. Add the block_year
    # label to each identified crown
    tree_shp <- crown_metrics(
      las_combine, .stdtreemetrics, geom = "concave", concaveman = c(1, 0)) |> 
      mutate(block = x) |> 
      dplyr::relocate(block, 1) |>  
      dplyr::relocate(geometry, .after = last_col())
    plan(sequential)
    
    # Write the crown POLYGON to a geopackage file
    st_write(tree_shp, file.path(tree_shp_dir, paste0(x, "_crowns.gpkg")), 
             quiet = TRUE, delete_dsn = TRUE)
    
    # Also generate individual POLYGON bounding boxes for each tree (it is a much
    # smaller file size and can be more easily loaded into a GIS for 
    # visualization). Write these to a geopackage.
    tree_box <- st_sf(data.table::rbindlist(lapply(1:nrow(tree_shp), function(i) {
      xx <- tree_shp[i, ]
      st_geometry(xx) <- st_as_sfc(st_bbox(st_geometry(xx)))
      return(xx)
    })))
    
    st_write(tree_box, file.path(tree_shp_dir, paste0(x, "_bbox.gpkg")), 
             quiet = TRUE, delete_dsn = TRUE)
    
    return(tree_shp)
  } else {
    plan(sequential)
    return(ttops)
  }
}) |> setNames(blocks)

```

The above returned object, `trees`, is a list item that contains either tree POINT or POLYGON data for the treetops or tree crowns, respectively. The data has been written to a geopackage file, and the purpose of this script has been fulfilled: trees have been successfully detected, segmented, or both.

The next step is to identify trees within reserves, strip edges, controls, or boundaries for use in future statistical practices. This will be completed in the next file, `03_tree_models`.
