---
title: "03_tree_models"
author: "Matthew Coghill"
date: "2023-01-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We should no longer have to work with the lidR package here since everything we needed from it was completed in the previous 2 R markdown files. The job of this file will be as follows:

1.  Load the generated tree shapes from the previous file;
2.  Load in the Goudie shape showing strip and reserve ID's;
3.  Load in the tree height and DBH data from the field;
4.  Model tree DBH from the LiDAR using the existing field data. 3 models are created:

```{=html}
<!-- -->
```
i.  A model for the strip edge trees which uses height, strip width, and aspect data from the edge trees in the field data;
ii. A model for the reserve trees which uses height and reserve width data from the reserve trees in the field data; and,
iii. A model for the control and boundary trees which is only based on height data from all trees in the field data

```{=html}
<!-- -->
```
5.  Calculate the volume of each tree using the FAIBBase package function `treeVolCalculator`

First, we need to load in the packages to be used.

```{r Packages, warning=FALSE, message=FALSE}

ls <- c("tidyverse", "sf", "sfheaders", "FAIBBase", "readxl", "data.table")
invisible(lapply(ls, library, character.only = TRUE))
rm(ls)
setDTthreads(0L)

```

Before looping over each block/year, we need to do some shape manipulation. The shapes that are provided include the areas for what is a reserve, strip, control, or the shape of the entire block. The eventual goal will be to define what tree shapes are considered as the edge of a strip, what trees are in the reserves, which ones are in the controls, and which trees are beyond those in the "boundary".

To give a bit more leeway to the definition of a strip edge, I decided to move the strip edge lines into the strip by 2.4m on each side. Before doing that, sometimes the trees that were defined as "edges" were actually behind an actual edge tree, hence why I decided to move these lines into the strips. These new strip edges get converted to individual points separated by 1m, and then the trees that are closest to each point are identified as being an "edge" tree.

After moving these points closer to the middle of the strip, sometimes the closest points to, say, the Northern strip edge were from the Southern strip. This was not ideal; thus, I devised a way to make sure that when the edge classification is completed it is only considering points from the reserve that the strip is adjacent to. This was done by joining the reserve shape with a buffered line (2.4m). Now, that 2.4m extends on both sides of the line: 2.4m towards the reserve, and 2.4m further into the strip. This is why I chose 2.4m: Think of a 10m wide strip. If the strip edge on one side was moved 2.4m in, and then buffered again by 2.4m, that is a total of 4.8m into the strip on one side. Factoring in the other side, we have a total of 9.6m extra space for being able to determine what is an edge tree. If that size had been 2.5m, then the polygons would touch each other and merge together into one big blob which is not what I wanted at all, hence the 2.4m value given. Hope that makes a bit of sense! The coding is a bit weird to run, but it works well.

Below, we tackle all of that. We also read in the provided tree information from field data and do a bit of manipulation on that. Here is where we could also filter out forked trees as well if that is something that needs to be considered. The script is currently set up to take just the treetop information; however, a slight tweak to the file pattern for reading in the tree shapes would allow for the polygons to be loaded in as well.

```{r Shape manipulation}

# Get and set directories for inputs and outputs
shape_dir <- file.path("./00_Shapes")
tree_shp_dir <- file.path("./06_tree_shp")
out_dir <- file.path("./07_results")
dir.create(out_dir, showWarnings = FALSE)

# Read and manipulate field data, filter tree heights greater than/equal to 10m
# Adjusted width of reserve near 10m strips to be 21m instead of 20m for modelling
# Strip orientation data can be detected from the 'Trt' column; if the value ends
# with "-N", "-S", or "-EW", strip orientation is "EW", otherwise it's "NS"
field_data <- read_excel("./00_field_data/2022 Tree Measurements- clean.xlsx") |> 
  mutate(
    Width = case_when(
      substr(transect, 2, 2) == 2 & startsWith(transect, "T") ~ 10,
      substr(transect, 2, 2) == 3 & startsWith(transect, "T") ~ 15,
      substr(transect, 2, 2) == 4 & startsWith(transect, "T") ~ 20,
      substr(transect, 2, 2) == 2 & startsWith(transect, "R") ~ 21,
      substr(transect, 2, 2) == 3 & startsWith(transect, "R") ~ 30,
      substr(transect, 2, 2) == 4 & startsWith(transect, "R") ~ 40,
      TRUE ~ 0),
    Treatment = case_when(
      startsWith(transect, "T") ~ "Treatment",
      startsWith(transect, "R") ~ "Reserve",
      TRUE ~ "Control"),
    Strip_Orientation = case_when(
      Trt %in% c("10-N", "15-N", "20-N", "10-S", "15-S", "20-S",
                 "10-EW", "15-EW", "20-EW") ~ "EW",
      Trt %in% c("10-E", "15-E", "20-E", "10-W", "15-W", "20-W",
                 "10-NS", "15-NS", "20-NS") ~ "NS",
      TRUE ~ NA_character_),
    Aspect = substr(Trt, 4, 5)) |>
  mutate(
    Aspect = ifelse(startsWith(transect, "R"), NA, Aspect),
    Treatment_ID = case_when(
      startsWith(transect, "T") ~ paste0(Treatment, "_", Width, "m_", Aspect),
      startsWith(transect, "R") ~ paste0(
        Treatment, "_", floor(Width / 2), "m_", Strip_Orientation),
      TRUE ~ Treatment)) |> 
  dplyr::filter(Ht >= 10)

# Get list of files to read in and retrieve CRS information. This is where we
# can change whether the treetops are loaded or the tree polygons.
trees <- list.files(tree_shp_dir, pattern = "_treetops.gpkg$", full.names = TRUE)
trees_crs <- st_layers(trees[1])[["crs"]][[1]]

# Load Goudie polygons. Rename the Reserve treatments appending the strip 
# orientation to its name
goudie <- st_read(file.path(shape_dir, "Goudie.gpkg"), quiet = TRUE) |> 
  st_transform(trees_crs) |> 
  mutate(Treatment = if_else(
    startsWith(Treatment, "Reserve"), paste0(Treatment, "_", Strip_Orientation),
    Treatment))

# Read in the strip edge lines. Rename the strip treatments appending the aspect
# of the strip to its name, and change name to ID to match other geometries. 
# Join the edges with the reserve data that each strip is adjacent to.
reserves <- dplyr::filter(goudie, startsWith(Treatment, "Reserve")) |> 
  select(-Block)
strip_edges <- st_read(file.path(shape_dir, "Goudie_Strip_Edges.gpkg"), 
                       quiet = TRUE) |> 
  st_transform(trees_crs) |> 
  mutate(Treatment = paste0(Treatment, "_", Aspect)) |> 
  rename(ID = StripID) |> 
  st_join(reserves, join = st_is_within_distance, dist = 1, suffix = c("", "_Adjacent")) |> 
  st_set_agr("constant") 

# Adjust the lines coordinates to be a bit more inside by 2.4m on each side.
# Get the geometry attribute name to properly do this (sometimes it's not called
# "geom" or "geometry" depending on the PC used)
g <- attr(strip_edges, "sf_column")
strip_edges[[g]][strip_edges$Aspect == "S"] <- strip_edges[[g]][
  strip_edges$Aspect == "S"] - c(0, 2.4)

strip_edges[[g]][strip_edges$Aspect == "N"] <- strip_edges[[g]][
  strip_edges$Aspect == "N"] + c(0, 2.4)

strip_edges[[g]][strip_edges$Aspect == "W"] <- strip_edges[[g]][
  strip_edges$Aspect == "W"] - c(2.4, 0)

strip_edges[[g]][strip_edges$Aspect == "E"] <- strip_edges[[g]][
  strip_edges$Aspect == "E"] + c(2.4, 0)

# Filter Goudie object for strips, reserves, controls, and boundaries
controls <- dplyr::filter(goudie, startsWith(Treatment, "C"))
strips <- dplyr::filter(goudie, startsWith(Treatment, "Treatment"))
reserves <- dplyr::filter(goudie, startsWith(Treatment, "Reserve")) |> 
  st_set_agr("constant")
bounds <- dplyr::filter(goudie, startsWith(Treatment, "Boundary")) |> 
  st_set_agr("constant")

# Need to create reserve shape that is buffered into the strip to identify
# trees that are "behind" the edge trees:
# Step 1) Union the reserve shape with a buffered strip edge line
reserves_buf <- reserves |> 
  st_union(st_geometry(st_buffer(strip_edges, 2.4, endCapStyle = "FLAT"))) |> 
  st_union() |> 
  st_cast("POLYGON") |> 
  st_as_sf() |> 
  st_join(reserves, left = FALSE)

# Step 2) There are some duplicate elements from the joining function. Deal with
# those so that there is only one element per reserve shape. Arrange table in a 
# sequence that has duplicated elements directly below each other
reserves_dup <- reserves_buf |> 
  group_by(Block, ID, Treatment, Strip_Orientation) |> 
  dplyr::filter(n() > 1) |> 
  ungroup() |> 
  arrange(Strip_Orientation, Treatment, ID, Block)

reserves_nodup <- reserves_buf |> 
  group_by(Block, ID, Treatment, Strip_Orientation) |> 
  dplyr::filter(n() == 1) |> 
  ungroup() |> 
  st_cast("MULTIPOLYGON")

# Apply new geometries to only odd row numbers
iter <- 1:nrow(reserves_dup)
iter <- iter[!(iter %% 2 == 0)]

reserves_dup_fix <- do.call(rbind, lapply(iter, function(x) {
  i <- reserves_dup[c(x, x + 1), ]
  j <- st_union(i)
  k <- i[1, ]
  st_geometry(k) <- j
  return(k)
}))

# Step 3) Merge the non-duplicated rows and fixed duplicated rows back together
reserves_buf <- rbind(reserves_nodup, reserves_dup_fix) %>%
  rename(!!attr(controls, "sf_column") := attr(., "sf_column")) |> 
  st_set_agr("constant")

# Create a shape of the forested area that is not a strip, reserve, or control 
# within the boundaries of each block.
bound_areas <- rbind(reserves_buf, controls, strips) |> 
  st_union() |> 
  sf_remove_holes() |> 
  st_sym_difference(st_combine(bounds)) |> 
  st_as_sf() |> 
  st_intersection(bounds) |> 
  st_collection_extract("POLYGON") |> 
  group_by(ID, Treatment, Strip_Orientation) |> 
  summarise(.groups = "drop") |> 
  st_set_agr("constant")

# Add in boundary information to the strip edge lines
strip_edges_add <- strip_edges %>% 
  dplyr::filter(rowSums(is.na(.)) > 0) |> 
  select(Block, ID, Aspect, Treatment) |> 
  st_join(bound_areas, join = st_nearest_feature, dist = 2, suffix = c("", "_Adjacent"))

strip_edges <- strip_edges %>% 
  dplyr::filter(!rowSums(is.na(.)) > 0) |> 
  rbind(strip_edges_add)

# To identify trees on the edge of a strip, we will convert the strip_edges 
# object to a series of points by sampling along the line at 1m intervals. We 
# will then find the nearest tree to each of the points, and each unique tree
# that is returned will be classified as an edge tree.
strip_sample <- st_cast(strip_edges, "LINESTRING", warn = FALSE) |> 
  st_line_sample(density = 1) |> 
  st_cast("POINT") |>
  st_as_sf() |> 
  st_intersection(st_set_agr(st_buffer(strip_edges, 2.5), "constant"))

```

The `lapply()` loop below completes the following actions:

1.  Loads in the tree shapes from the previous script;
2.  Roughly identifies reserve trees
3.  Accurately identify the edge trees, then the reserve trees, then trees within the control area, then any trees that fell within the thinned strips, and finally trees that were beyond the Goudie sampling area
4.  Create an ANOVA model for predicting tree DBH from tree height within treatments using the field data
5.  Calculate the volume of each individual tree given that we are looking at lodgepole pine (PL) in the Montane Spruce BGC zone (MS) by using tree heights and DBH's

Once the loop completes, it returns a single data frame with information on each individual tree. We can then use that data in a statistical manner to understand the differences in tree parameters with respect to the orientation of the strips, their associated strip widths, and their aspects as well.

```{r Model tree DBH and volume}

# Get the DBH and volume of each tree in a tidy output
tree_measurements <- do.call(rbind, lapply(trees, function(x) {
  print(paste("Analyzing DBH and volumes for", x))
  
  # Read in the trees, filter the Goudie object to only include strips from the
  # same block as the trees being loaded in
  tree_shp <- st_read(x, quiet = TRUE) |> 
    rename(block_year = block) |> 
    dplyr::filter(Z >= 12) |> 
    st_set_agr("constant")
  goudie_flt <- dplyr::filter(goudie, startsWith(unique(tree_shp$block), Block))
  block_id <- unique(goudie_flt$Block)
  
  # Filter strip sampling points to only contain the data from that block
  strip_sample_flt <- dplyr::filter(strip_sample, Block == block_id)
  strips_flt <- dplyr::filter(strips, Block == block_id)
  strip_edge_flt <- dplyr::filter(strip_edges, Block == block_id)
  
  # Get the unique treatment ID's from this block, and filter the field data
  # to match those treatments.
  tree_areas <- c(
    unique(goudie_flt$Treatment[startsWith(goudie_flt$Treatment, "Reserve")]),
    unique(strip_edge_flt$Treatment), "Control", "Boundary")
  
  field_flt <- dplyr::filter(field_data, Treatment_ID %in% tree_areas) |> 
    mutate(Treatment = as.factor(Treatment),
           Aspect = as.factor(Aspect),
           Width = as.factor(Width))
  
  # First, create rough reserve identity for each tree
  tree_ids_res <- tree_shp[unlist(st_intersects(reserves_buf, tree_shp)), ] |> 
    st_join(reserves_buf, join = st_nearest_feature)
  
  # Detect duplicate features and handle them if present
  tree_ids_res_nodup <- st_drop_geometry(tree_ids_res) |> distinct()
  
  if(nrow(tree_ids_res) != nrow(tree_ids_res_nodup)) {
    g <- attr(tree_ids_res, "sf_column")
    rowids <- which(rownames(tree_ids_res) %in% rownames(tree_ids_res_nodup))
    st_geometry(tree_ids_res_nodup) <- tree_ids_res[[g]][rowids]
    tree_ids_res_nodup <- st_set_geometry(tree_ids_res_nodup, g)
  } else tree_ids_res_nodup <- tree_ids_res

  # Next, identify trees that are along the boundaries of the block
  tree_ids_bounds <- tree_shp[unlist(st_intersects(bound_areas, tree_shp)), ] |> 
    st_join(bound_areas, join = st_nearest_feature) |> 
    mutate(Block = block_id) |> 
    dplyr::filter(!treeID %in% tree_ids_res_nodup$treeID)
  
  tree_ids <- rbind(tree_ids_res_nodup, tree_ids_bounds)
  
  # Next, determine if a tree is along the edge going one strip at a time to 
  # ensure that trees are only being detected along the adjacent reserve
  edge_join <- do.call(rbind, lapply(1:nrow(strip_edge_flt), function(i) {
    strip_edge_i <- strip_edge_flt[i, ]
    strip_sample_i <- strip_sample_flt |> 
      dplyr::filter(Block == strip_edge_i$Block,
                    ID == strip_edge_i$ID,
                    Treatment == strip_edge_i$Treatment,
                    Aspect == strip_edge_i$Aspect)
    edge_bound <- st_geometry(st_buffer(strip_edge_i, 50, endCapStyle = "FLAT"))
    tree_select <- tree_ids |> 
      dplyr::filter(ID == strip_edge_i$ID_Adjacent,
                    Treatment == strip_edge_i$Treatment_Adjacent) |> 
      mutate(Aspect = strip_edge_i$Aspect,
             Treatment = strip_edge_i$Treatment,
             ID = strip_edge_i$ID)
    output <- tree_select[unique(st_nearest_feature(strip_sample_i, tree_select)), ]
    output <- output[unlist(st_intersects(edge_bound, output)), ] |> 
      st_set_agr("constant")
    return(output)
  }))
  
  # DEAL WITH DUPLICATED EDGE TREES - decision for which strip edge they would
  # belong to if their polygon boundaries extend into two reserves (the one they
  # are closest to)
  edge_dup <- edge_join |> 
    dplyr::filter(treeID %in% edge_join$treeID[duplicated(edge_join$treeID)])
  
  if(nrow(edge_dup) > 0) {
    tree_edge_fix <- do.call(rbind, lapply(unique(edge_dup$treeID), function(i) {
      tree_edge_flt <- edge_dup[edge_dup$treeID == i, ]
      edge_choice <- strip_edge_flt[
        which.min(st_distance(strip_edge_flt, tree_edge_flt)[, 1]), ]
        
      output <- tree_edge_flt |> 
        dplyr::filter(ID == edge_choice$ID, Aspect == edge_choice$Aspect)
    }))
    rowids <- which(rownames(edge_join) %in% rownames(tree_edge_fix))
    edge_join <- rbind(
      edge_join[!edge_join$treeID  %in% unique(edge_dup$treeID), ],
      edge_join[rowids, ])
  }
  
  # Remove identified edge trees from the main pool of trees. Now, identify 
  # trees fully within the reserves
  trees_remain <- dplyr::filter(tree_shp, !treeID %in% edge_join$treeID)
  reserve_trees <- trees_remain[unlist(st_intersects(reserves_buf, trees_remain)), ]
  reserve_join <- st_join(reserve_trees, reserves_buf, join = st_nearest_feature)
  
  # DEAL WITH DUPLICATED RESERVE TREES - decision for which reserve they would
  # belong to if their polygon boundaries extend into two reserves (the one they
  # are closest to)
  reserve_dup <- reserve_join |> 
    dplyr::filter(treeID %in% reserve_join$treeID[duplicated(reserve_join$treeID)])
  
  if(nrow(reserve_dup) > 0) {
    tree_res_fix <- do.call(rbind, lapply(unique(reserve_dup$treeID), function(i) {
      tree_res_flt <- reserve_dup[reserve_dup$treeID == i, ][1, ]
      tree_res_flt_fix <- tree_res_flt |> 
        st_set_agr("constant") |> 
        st_intersection(reserves_buf) %>%
        slice_max(st_area(.)) |> 
        mutate(ID = ID.1) |> 
        select(-c(ID.1, Block.1, Treatment.1, Strip_Orientation.1)) |> 
        st_set_geometry(st_geometry(tree_res_flt))
    }))
    rowids <- which(rownames(reserve_join) %in% rownames(tree_res_fix))
    reserve_join <- rbind(
      reserve_join[!reserve_join$treeID  %in% unique(reserve_dup$treeID), ],
      reserve_join[rowids, ])
  }
  
  # Remove identified reserve trees. Now, identify trees in the control blocks
  trees_remain <- dplyr::filter(trees_remain, !treeID %in% reserve_join$treeID)
  control_trees <- trees_remain[unlist(st_intersects(controls, trees_remain)), ]
  control_join <- st_join(control_trees, controls) |> 
    mutate(Strip_Orientation = NA)
  
  # Remove identified control trees. Now, identify trees fully within the strips
  # and classify them as reserve trees
  trees_remain <- dplyr::filter(trees_remain, !treeID %in% control_join$treeID)
  strip_trees <- trees_remain[unlist(st_intersects(strips_flt, trees_remain)), ]
  strip_join <- st_join(strip_trees, strips)
  
  # Remove trees outside of controls, strips, and reserves. Now, identify
  # remaining trees
  trees_remain <- dplyr::filter(trees_remain, !treeID %in% strip_join$treeID) |>
    mutate(Block = block_id, ID = "B", Treatment = "Boundary", 
           Strip_Orientation = NA)
  
  # Merge everything back together, ensure that only trees from the respective
  # block are included or modelling will fail
  tree_shp_merge <- rbindlist(
    list(edge_join, strip_join, reserve_join, control_join, trees_remain), 
    fill = TRUE) |> 
    st_sf() |> 
    rename(Treatment_ID = Treatment) |> 
    mutate(
      Aspect = as.factor(Aspect),
      Treatment = as.factor(gsub("_.*", "", Treatment_ID)),
      Width = as.factor(case_when(
        substr(ID, 2, 2) == 2 & Treatment == "Treatment" ~ 10,
        substr(ID, 2, 2) == 3 & Treatment == "Treatment" ~ 15,
        substr(ID, 2, 2) == 4 & Treatment == "Treatment" ~ 20,
        substr(ID, 2, 2) == 2 & Treatment == "Reserve" ~ 21,
        substr(ID, 2, 2) == 3 & Treatment == "Reserve" ~ 30,
        substr(ID, 2, 2) == 4 & Treatment == "Reserve" ~ 40,
        TRUE ~ 0)))
  
  # Create models of DBH for edge trees, reserve trees, and control/boundary 
  # trees. Sometimes, trees are detected within the middle of a strip. Those are
  # not classified as an edge or a reserve tree, rather they fall in the same
  # category as a control or boundary tree for the purposes of modelling. Those
  # trees also don't have an aspect defined so that they are treated as their
  # own entity. 
  model_trt <- lm(DBH ~ Ht * Width * Aspect, data = subset(field_flt, Treatment == "Treatment"))
  model_res <- lm(DBH ~ Ht * Width, data = subset(field_flt, Treatment == "Reserve"))
  model_ctl <- lm(DBH ~ Ht, data = field_flt)
  
  tree_shp_trt <- tree_shp_merge |> 
    dplyr::filter(Treatment == "Treatment", Treatment_ID %in% tree_areas) |> 
    mutate(DBH = predict(model_trt, newdata = data.frame(
      Ht = Z, Width = Width, Aspect = Aspect)))
  tree_shp_res <- tree_shp_merge |> 
    dplyr::filter(Treatment == "Reserve", Treatment_ID %in% tree_areas) |> 
    mutate(DBH = predict(model_res, newdata = data.frame(Ht = Z, Width = Width)))
  tree_shp_ctl <- tree_shp_merge |> 
    dplyr::filter(!treeID %in% c(tree_shp_trt$treeID, tree_shp_res$treeID)) |> 
    mutate(DBH = predict(model_ctl, newdata = data.frame(Ht = Z)))
  
  # Combine that data back together
  tree_dbh <- rbind(tree_shp_trt, tree_shp_res, tree_shp_ctl) 
  
  # Estimate the volume of each tree in the LiDAR data
  tree_shp_merge_vols <- FAIBBase::treeVolCalculator(
    FIZorBEC = "MS", species = "PL", height = tree_dbh$Z, DBH = tree_dbh$DBH)
  
  # Create output table combining volume and height data
  tree_vols <- tree_dbh |> 
    mutate(whole_tree_vol = tree_shp_merge_vols$VOL_WSV,
           total_merch_vol = tree_shp_merge_vols$VOL_BELOW_UTOP,
           stump_vol = tree_shp_merge_vols$VOL_STUMP,
           non_merch_vol = tree_shp_merge_vols$VOL_ABOVE_UTOP) |> 
    dplyr::relocate(attr(tree_dbh, "sf_column"), .after = last_col())
  
  return(tree_vols)
}))

# Summarize the tree measurement data
tree_shp_data <- st_drop_geometry(tree_measurements) |> 
  group_by(Treatment_ID, block_year) |> 
  summarise(mean_height = mean(Z), 
            mean_volume = mean(total_merch_vol), n = n(), 
            min_height = min(Z), max_height = max(Z), .groups = "drop")

# Calculate stem density. First, remove edge trees. Then, group variables and
# summarise the data to get the number of trees in each strip/reserve/control
# area. Join it back with the shapes of Goudie which contain info on the spatial
# area. Change any missing NA values to 0's since they represent 0 stems/ha
units::install_unit("stems")

stem_density <- st_drop_geometry(tree_measurements) |> 
  dplyr::filter(!startsWith(Treatment_ID, "Treatment_")) |> 
  group_by(Treatment_ID, Block, block_year, ID) |> 
  summarise(n = n(), .groups = "drop") |> 
  full_join(goudie, by = c("Treatment_ID" = "Treatment", "Block", "ID")) |> 
  dplyr::filter(!grepl("^T|^B", ID)) |> 
  st_sf() %>%
  dplyr::mutate(
    n = units::set_units(ifelse(is.na(n), 0, n), "stems"),
    area = units::set_units(st_area(.), "ha"), density = n / area) |> 
  dplyr::relocate(attr(tree_dbh, "sf_column"), .after = last_col())

# Summarise density data by treatment and orientation
stem_density_sum <- st_drop_geometry(stem_density) |> 
  group_by(block_year, Treatment_ID) |> 
  summarise(mean_density = mean(density), .groups = "drop")

# Write results
write.csv(st_drop_geometry(tree_measurements), file.path(
  out_dir, "tree_measurements.csv"), row.names = FALSE)
write.csv(st_drop_geometry(stem_density), file.path(
  out_dir, "stem_density_by_strip.csv"), row.names = FALSE)
write.csv(stem_density_sum, file.path(out_dir, "stem_density_summary.csv"),
          row.names = FALSE)

```

The data generated above can be used in multiple statistical analyses, it just depends on the questions you want to ask.
